Fast Sorting on Modern Computers

### 2005-08-26 docs/tech_report/elementary.tex
more changes due to davids minor edits

It should be possible to make more cache-conscious versions of these sorts.
Making a bubblesort which performs a bubble across an entire cache line would
be an example. However, these would only make a difference on large data sets,
when better sorts such as quicksort should be used. In addition, the complexity
of doing this means it may be easier to code a simple $O(NlogN)$ sort, which
would perform better for the same effort. Due to this, a cache-conscious
elementary sort is but an academic exercise.

>>>>

It should be possible to make more cache-conscious versions of these sorts.
Making a bubblesort which performs a bubble across an entire cache line would be
an example. Another would be a cache-conscious selection sort. It might switch
the direction of the search to reuse lines at the end of the array. A
cache-conscious bubblesort could also do something similar, which would make it
into shakersort. However, these would only make a difference on large data sets,
when better sorts such as quicksort should be used. In addition, the complexity
of improving elementary sorts means it may be easier to code a simple $O(NlogN)$
sort, which would perform better for the same effort. Due to this, a
cache-conscious elementary sort is but an academic exercise.

### 2005-08-26 docs/tech_report/heap.tex
fixed minor typo
more changes due to davids minor edits

A \n{binary tree} is a linked data structure where a node contains data and links
to two other nodes, which have the same structure. The first node is called the
\n{parent}, while the linked nodes are called \n{children}. One node is called the
\n{root}, and the tree branches out from this, with each level having twice as many
nodes as the previous level.

A heap is a type of binary tree with several important properties. Firstly, the
the data stored by the parent must always be greater than the data stored by its
children. Secondly, in a binary tree there is no requirement that the tree be full.
In a heap, it is important that every node's left child exists before its right
child. With this property, a heap can be stored in an array without any wasted
space, and the links are implied by a nodes position, rather than by a pointer
or reference. A heap done in the fashion is called an \n{implicit heap}.

@@

Floyd suggested an improvement to this, where the new root is sifted straight
down without regard for whether it is larger than its child. Typically, it will
end near the bottom, so sifting up from here should save comparisons. Sedgewick
notes speed improvements only when comparisons are move expensive than swaps,
such as in the case of sorting strings. For integers, this step should not be
done.

@@

LaMarca overcomes the problem by inserting a maximal key at the end of a heap if
there's an even number of nodes in the array. This gives a second child to the
last element of the heap. This is an effective technique for building the heap,
as the number of keys is static. When destroying the heap, however, the
maximal key must be added to every second operation. This can be avoided
relatively simply with no extra instructions. If the last key in the heap is
left in place as well as being sifted down the heap, it will provide a
sentinel for the sift-down operation. The key from the top of the heap is then
put into place at the end. This technique reduces instruction count by
approximately 10\%.

@@

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
2.58 mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions (about 36\%) due to the combination of the shallower heap
and the larger fan-out.

>>>>

A \n{binary tree} is a linked data structure where a node contains data and links
to two other nodes, which have the same structure. The first node is called the
\n{parent}, while the linked nodes are called \n{children}. There is one node
with no parents, called the \n{root}, from which the tree branches out, with
each level having twice as many nodes as the previous level.

A heap is a type of binary tree with several important properties. Firstly, the
data stored by the parent must always be greater than the data stored by its
children. Secondly, in a binary tree there is no requirement that the tree be
full.  In a heap, it is important that every node's left child exists before its
right child. With this property, a heap can be stored in an array without any
wasted space, and the links are implied by a nodes position, rather than by a
pointer or reference. A heap done in the fashion is called an \n{implicit heap}.

@@

Floyd suggested an improvement to this, where the new root is sifted straight
down without regard for whether it is larger than its child. Typically, it will
end near the bottom, so sifting up from here should save comparisons. Sedgewick
notes speed improvements only when comparisons are move expensive than swaps,
such as in the case of sorting strings. In our case, the comparisons are very
predictable moving down the heap, and so this step should not be done.

@@

LaMarca overcomes the problem by inserting a maximal key at the end of a heap if
there's an even number of nodes in the array. This gives a second child to the
last element of the heap. This is an effective technique for building the heap,
as the number of keys is static. When destroying the heap, however, the
maximal key must be added to every second operation. We found this can be
avoided with an innovative technique, with no extra instructions. If the last
key in the heap is left in place as well as being sifted down the heap, it will
provide a sentinel for the sift-down operation. The key from the top of the heap
is then put into place at the end. This technique reduces instruction count by
approximately 10\%.

@@

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
1.59 mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions (about 36\%) due to the combination of the shallower heap
and the larger fan-out.

### 2005-08-26 docs/tech_report/merge.tex
added a label
more changes due to davids minor edits

\subsection{Algorithm S}

@@

Overall, when the $k$-ary heap is used, there are approximately the same total
number of branch mispredictions as in the other mergesort algorithms. Although
the final $k$-way merge executes a lot more comparison branches, these heap
branches are quite predictables, as we saw in Section
\ref{heapsort-branch-prediction}. This means that using the technique to reduce
cache misses does not adversely affect branch prediction rates (it does,
however, still affect the instruction count).  The rate of branch misses is much
the same as that in heapsort, and progresses in the same manner (see Figure
\ref{memory-tuned heapsort curve}). Note that `\n{comparison 0}' is included for
illustration purposes only. It is guaranteed to be taken every time, and would
be optimized out by the compiler.

>>>>

\subsection{Algorithm S}
\label{Algorithm S}

@@

Overall, when the $k$-ary heap is used, there are approximately the same total
number of branch mispredictions as in the other mergesort algorithms. Although
the final $k$-way merge executes a lot more comparison branches, these heap
branches are quite predictable, as we saw in Section
\ref{heapsort-branch-prediction}. This means that using the technique to reduce
cache misses does not adversely affect branch prediction rates (it does,
however, still affect the instruction count).  The rate of branch misses is much
the same as that in heapsort, and progresses in the same manner (see Figure
\ref{memory-tuned heapsort curve}). Note that `\n{comparison 0}' is included for
illustration purposes only. It is a comparison between two values, one of which
has just been assigned to the other, and which are therefore equal. It is
guaranteed to be taken every time, and would be optimized out by the compiler.

### 2005-08-26 docs/tech_report/quick.tex
more changes due to davids minor edits

LaMarca uses the optimized quicksort described by Sedgewick for his base
quicksort. This is an iterative quicksort which uses an explicit stack. In the
worst case, this stack must be the same size as the array, and must be
initialised at the start of the algorithm. To reduce the size of the stack,
Sedgewick proposes sorting the smallest partition first. This optimization
reduces the maximum size of the stack to $O(logN)$. In practice, two \cc{int}s,
the left and the right index, are required for each partition.

Sedgewick's second improvement is the use of an elementary sort for small
sub-files. Once a partition is smaller than an arbitrary threshold, insertion
sort is used to sort the partition. Sedgewick states that roughly the same
results are obtained for thresholds between 5 and 25. An additional reduction in
instruction count can be had by doing the insertion sort over the entire array
at the end, rather than across each small array as soon as the threshold is
reached. At the end of the partition sorting, LaMarca places a sentinel at the
end of the array and uses an insertion sort, eliminating the bounds check from
the end. In this implementation, the sentinel is placed at the front, and the
smallest of the first \cc{THRESHOLD} keys is used as a sentinel, in keeping
with our policy of not exceeding the bounds of the array.

@@

LaMarca notes that quicksort is already very cache efficient as it has excellent
spatial and temporal locality. Once it no longer fits in the cache, however, the
number of misses should increase significantly. This is especially true in the
case of the base quicksort, which does an insertion sort across then entire
array at the end.

@@

\label{quicksort entropy}
When we originally looked at quicksort, we expected that the \n{i} and \n{j}
branches in quicksort would resolve in each direction. In fact, this is not the
case. The main reason is that the chosen pivot is almost never the true median
of the section of the array to be partitioned. Thus, the \n{i} and \n{j}
branches are almost always biased, which makes them predictable. According to
Knuth, if the pivot element is chosen randomly, then the \n{i} and \n{j}
branches will be taken (i.e. the \n{i} or \n{j} loop will execute) twice as
often as they are not taken (i.e. the loop exits).

@@

For example, consider the case where the pivot element is larger than the true
median. In this case, the final resting place of the pivot will be towards the
right-hand end of the region to be partitioned. Thus, the \n{i} loop will
execute a large number of times, and thus the \n{i} loop will be biased towards
being taken. The \n{j} loop, on the other hand, will be biased toward not-taken,
because the average element is likely to be smaller than the pivot. Therefore,
the \n{j} loop will execute only a small number of times. Thus, we see that in a
particular partition, we do not normally get an equal number of executions of
the \n{i} and \n{j} loops. Instead we get a large number of executions in one
loop (which makes its branch relatively predictable), whereas the other loop
often does not enter at all (which makes its branch also predictable, although
executed a much smaller number of times). Thus, although overall the \n{i} and
\n{j} branches are taken twice as often as not-taken, on a particular partition
one will usually be biased in the not-taken direction, and as a result will be
executed a relatively small number of times.

@@

\section{Future Work}
The threshold in quicksort is the point at which a simpler sort is used to sort
the partition. A larger threshold will results in fewer checks for a sentinel. A
smaller threshold means that the instruction count due to the insertion sort
will be lower, but more partition steps may be taken. A smaller threshold also
makes it more likely that the keys to be sorted all fit in a single cache
block. However, the partition is likely to occur across a cache block boundary,
though this is not necessarily a bad thing; the next partition to be sorted will
be adjacent to the current block, and will already be in memory. Observing how
results, especially instruction count and branch misses, vary with changes 
to the threshold may be interesting.

>>>>

LaMarca uses the optimized quicksort described by Sedgewick in
\cite{Sedgewick02} for his base quicksort. This is an iterative quicksort which
uses an explicit stack. In the worst case, this stack must be the same size as
the array, and must be initialised at the start of the algorithm. To reduce the
size of the stack, Sedgewick proposes sorting the smallest partition first. This
optimization reduces the maximum size of the stack to $O(logN)$. In practice,
two \cc{int}s, the left and the right index, are required for each partition.

Sedgewick's second improvement is the use of an elementary sort for small
sub-files. Once a partition is smaller than an arbitrary threshold, insertion
sort is used to sort the partition. Sedgewick states that roughly the same
results are obtained for thresholds between five and 25. An additional reduction
in instruction count can be had by doing the insertion sort over the entire
array at the end, rather than across each small array as soon as the threshold
is reached. At the end of the partition sorting, LaMarca places a sentinel at
the end of the array and uses an insertion sort, eliminating the bounds check
from the end. In this implementation, the sentinel is placed at the front, and
the smallest of the first \cc{THRESHOLD} keys is used as a sentinel, in keeping
with our policy of not exceeding the bounds of the array.

@@

LaMarca notes that quicksort is already very cache efficient\footnote{We
speculate wildly at this point and suggest that since an average of 25\% of
computing time is spent sorting (see \cite{TODO}), the development of caches has
been subconsicously affected by the cache performance of quicksort.}, as it has
excellent spatial and temporal locality. Once it no longer fits in the cache,
however, the number of misses should increase significantly. This is especially
true in the case of the base quicksort, which does an insertion sort across then
entire array at the end.

@@

\label{quicksort entropy}
When we originally looked at quicksort, we expected that the \n{i} and \n{j}
branches in quicksort would resolve in each direction an equal number of times.
In fact, this is not the case. The main reason is that the chosen pivot is
almost never the true median of the section of the array to be partitioned.
Thus, the \n{i} and \n{j} branches are almost always biased, which makes them
predictable. According to Knuth, if the pivot element is chosen randomly, then
the \n{i} and \n{j} branches will be taken (i.e. the \n{i} or \n{j} loop will
execute) twice as often as they are not taken (i.e. the loop exits).

@@

For example, consider the case where the pivot element is larger than the true
median. In this case, the final resting place of the pivot will be towards the
right-hand end of the region to be partitioned. Thus, the \n{i} loop will
execute a large number of times, and thus the \n{i} branch will be biased
towards being taken. The \n{j} branch, on the other hand, will be biased toward
not-taken, because the average element is likely to be smaller than the pivot.
Therefore, the \n{j} loop will execute only a small number of times. Thus, we
see that in a particular partition, we do not normally get an equal number of
executions of the \n{i} and \n{j} loops. Instead we get a large number of
executions in one loop (which makes its branch relatively predictable), whereas
the other loop often does not enter at all (which makes its branch also
predictable, although executed a much smaller number of times). Thus, although
overall the \n{i} and \n{j} branches are taken twice as often as not-taken, on a
particular partition one will usually be biased in the not-taken direction, and
as a result will be executed a relatively small number of times.

@@

\section{Future Work}
The threshold in quicksort is the point at which a simpler sort is used to sort
the partition. A larger threshold will results in fewer checks for a sentinel. A
smaller threshold means that the instruction count due to the insertion sort
will be lower, but more partition steps may be taken. A smaller threshold also
makes it more likely that the keys to be sorted all fit in a single cache
block. However, the partition is likely to occur across a cache block boundary,
though this is not necessarily a bad thing; the next partition to be sorted will
be adjacent to the current block, and will already be in memory. Observing how
results, especially instruction count and branch misses, vary with changes 
to the threshold may be interesting. In particular, we expect that a larger
threshold would be beneficial, as insertion sort causes far fewer branch
mispredictions than quicksort.
